The Soulgem Paper

### 2011-09-22 soulgem.tex
Findings - description of reasons for clustering of expectations.

We proposed that expectations are clustered in roles based on a combination of institutional, historical, structural, and personal factors. Our empirical evidence provided support for this proposition, as detailed below.

\emph{Institutional factors:} The larger culture plays an important part in shaping the expectations of members of software organizations. For example, when Gallium hired one of its Product Managers, he had significant domain expertise, but no software development experience. He initially struggled to satisfy the expectations placed on him. (...)

\emph{Historical factors:}

\emph{Structural factors:}

\emph{Personal factors:}

>>>>

We proposed that expectations are clustered in roles based on a combination of institutional, historical, structural, and personal factors. Our empirical evidence provided support for this proposition, as detailed below. For space reasons we cannot summarize more than one example of each kind.

\emph{Institutional factors:} Although the software industry's culture still leaves plenty of room for role variability, it contributes to shape role definitions. We have several examples of this, but the widest in scope is when Gallium, after struggling for a few years as a self-funded start-up, acquired a considerable amount of venture capital, it quickly expanded and began distributing its labour in predictable ways: it created several groups (Product Management, Quality Assurance, Sales, Customer Support, Human Resources) and filled them up with newcomers with the goal of taking care of that aspect of the business. Before this influx, the division of labour was more flexible and organic: most organization members did a bit of everything. After the influx, role definitions shifted towards a more stereotypical differentiation of roles and groups.

\emph{Historical factors:} Previous events in the history of the organization shape its current clustering of expectations, in the same way that alterations early in the life of other organic systems shape their later structure. Some developers that have been part of the organization for a long time are not just developers: they are also mentors, historians, and political players, despite what the organizational chart says about them. This affects the expectations that people place on them. During the Gallium downsizing, some people were demoted to keep the company afloat. One of them was the previous head of the QA department (which was dispersed) and, later, Program Manager. At the time of our observations he was a QA Analyst, yet he was still involved (and expected to be involved) in many strategic activities that do not match the institutional view of what a QA Analyst is supposed to do.

\emph{Structural factors:} A person's position in the organizational structure gives her a certain power and awareness that others may not have. Her structural position leads others to expect from her behaviours relating to its characteristics: she may be expected to play a bridging, directing, or even confrontational role. Many of the expectations placed on the head of the Operations group at Cobalt spawn from his structural position. He is expected to provide news and updates of the operational status of Cobalt's services, he is expected to deal with and shelter the Software group from the System Administration group of the Institute that Cobalt is a part of, and he is expected to control the gateway that determines whether a product is ready to be released or not, that is, to assess the operational quality of the product. These expectations largely spawn from his position in the organization; it would be hard for someone without his position to satisfy them.

\emph{Personal factors:} Finally, a person's background, preferences, and personality shapes the expectations she receives from others. Perhaps the person's background could be deficient for the fulfillment of his initial role, and the organization adapts by expecting different or lesser things from him. Alternatively, a person's traits (curiosity, proactivity, a talent) may lead him to take on responsibilities that would not otherwise be placed on him given institutional, historical, or structural factors. For example, the Project Administrator at Cobalt was hired to perform mainly clerical tasks, but his own preference led him to gradually take more and greater responsibilities. The organization learned to be dependent on such behaviour. When the Project Administrator temporarily left the group to fulfill a position elsewhere in the Institute, Cobalt brought in a temporary replacement, and its projects suffered quickly and significantly, as the replacement did not bring the initiative of the original Project Administrator, but rather focused on the expectations that were officially placed on him. We should also note that at the time of our observations the Project Administrator position was undergoing an official redefinition in order to match real expectations better. In this way, personal factors were leading to historical and structural changes in the organization, so that future enactors of the role played by the Project Administrator would be affected by them.
